% ---- Abstract in English ----
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract (English)}
\markboth{ABSTRACT}{ABSTRACT}

Automatic short-answer grading (ASAG) has advanced rapidly for high-resource languages, but
Khmer, the official language of Cambodia, spoken by more than 17 million people, has, to our
knowledge, no prior published ASAG benchmark. This thesis introduces the first reproducible Khmer
ASAG benchmark: a corpus of 1{,}184 teacher-graded short answers across four secondary-school
subjects, three curated cleaning variants, and a unified pipeline that compares four families of
grading model: a classical model (TF-IDF features with a support-vector regressor), a
character-level BiLSTM with attention, multilingual Transformer encoders (mBERT, XLM-R, and
GTE-multilingual), and a large language model fine-tuned with QLoRA (Qwen 3.5 4B). Because the thesis
foregrounds \emph{explainability}, it additionally conducts a cross-family explainability study that
\emph{measures} explanation faithfulness (ERASER comprehensiveness and sufficiency, and the
area-over-the-perturbation-curve, AOPC) and reference-answer plausibility, rather than merely
displaying word-importance highlights.

The study has three main findings. First, the four families are \emph{comparable} on Quadratic
Weighted Kappa (QWK) (0.795, 0.845, 0.820, and 0.843, respectively, a narrow 0.05 band), while the
fine-tuned LLM is the clear best on the deployment metrics that matter in a classroom (66\% exact
integer-score match and 83\% agreement within one point). Second, \emph{Leave-One-Out (LOO) word-attribution} (text highlighting) explanations are
\emph{reliably} faithful for every model family, as verified by ERASER comprehensiveness and
sufficiency metrics, making LOO the unified, model-agnostic explanation method across all four pillars. Third, under a question-held-out evaluation the classical model's QWK collapses from
0.76 to 0.35, showing that small-question-pool ASAG scores are heavily inflated by \emph{question
leakage}. The explainability contribution is realized in a live, teacher-facing web prototype that
returns a score, a faithful word-importance explanation, and written feedback. All code, data,
random seeds, and figures are released for reproducibility.

\vspace{0.4cm}
\noindent\textbf{Keywords:} Automatic Short-Answer Grading; Explainable AI; Faithfulness; Khmer
Language Processing; Low-Resource NLP; Transformers; Quadratic Weighted Kappa.
\clearpage
