\chapter{Introduction}
\label{ch:intro}

\section{Background to the Study}
Assessment is a core part of education. It measures what students have learned, guides teaching, and
gives learners feedback on how to improve. Much of this assessment uses \emph{short-answer questions}:
a student writes one to a few sentences, which are then judged for \emph{content correctness} against
a reference answer. Short-answer questions sit between multiple-choice items (easy to mark but mostly
testing recall) and essays (rich but slow to grade). They are common in school exams, and grading
them by hand is slow and not always consistent, which limits how often teachers can give detailed
feedback.

\emph{Automatic Short-Answer Grading} (ASAG) studies how to score these answers with a computer. The
field has moved from early similarity-based systems to deep neural networks, and most recently to
Transformer-based and large language models~\citep{sung2022survey,devlin2019bert}. These advances have
improved ASAG for high-resource languages such as English, Arabic, and Chinese.

Those benefits have mostly \emph{not} reached low-resource languages. \textbf{Khmer}
(\km{ភាសាខ្មែរ}), the official language of Cambodia and spoken by more than 17 million people, is
still low-resource in NLP. The Khmer script is also hard to process:

\begin{itemize}[leftmargin=1.4em]
  \item it has \textbf{no spaces between words}, so even splitting a sentence into words is a
  machine-learning task in itself;
  \item it stacks subscript consonants (\km{ជើង}) and reorders vowels inside a syllable;
  \item it carries invisible formatting characters that survive naive cleaning.
\end{itemize}

Multilingual models see little Khmer during pre-training, and existing Khmer NLP work covers
tokenisation, part-of-speech tagging, and search, but not grading~\citep{buoy2022khmer,khmernltk2020}.
At the same time, AI in education is a growing priority for Cambodia~\citep{huot2025cambodia}, which
makes a reliable, transparent Khmer grading tool both timely and useful.

A grade is also a decision a teacher must be able to stand behind. A score with no justification is
hard to trust and harder to defend to a student, so a grading tool for a real classroom needs to do
more than output a number: it has to show \emph{why}. This study therefore treats a grading model as
something to be judged on three axes at once: how well it agrees with the teacher (accuracy), how often
it recovers the exact mark a teacher would write (deployment quality), and whether the explanation it
offers actually reflects how it decided (explainability). A model can be strong on one axis and weak on
another, so we measure all three and report them side by side rather than collapsing them into a single
ranking.

\section{Problem Statement}
We want to grade Khmer short answers automatically and to \emph{honestly} describe what is achievable
on a real classroom corpus. Three properties of the data make this hard:

\begin{itemize}[leftmargin=1.4em]
  \item \textbf{No benchmark and no baselines.} There is no published Khmer ASAG dataset or system, so
  we must build both the benchmark and the reference results.
  \item \textbf{Different scoring scales.} Each question has its own maximum score, from the set
  $\{5,6,7,8,10,12,15,20\}$. A raw score of ``8'' means different things on different questions, so the
  scores must be normalised.
  \item \textbf{A small, single-grader corpus.} The corpus has 1{,}184 answers graded by \emph{one}
  teacher, so there is limited data, no second grader to measure a human ceiling, and some real
  marking noise to confront.
\end{itemize}

A further problem, often blurred in ASAG papers, is that two different questions are usually reported
together: \emph{research quality} (how well the model's grades agree with the teacher, measured by
QWK) and \emph{deployment quality} (whether the model predicts the \emph{exact} mark in points). We
keep the two apart, because a model can be good at one and weak at the other. A model whose grades
\emph{correlate} well with the teacher's can still be the wrong tool for a classroom if it rarely hits
the exact mark, and a teacher who has to re-check every score gains little.

To make the task concrete, consider one real (translated) History item from the corpus. The
\textbf{question} asks which factors caused Siam to break the citadel of Longvek; the
\textbf{reference answer} lists several causes (the king's poor judgement, the loss of unity, and so
on) and is worth a \textbf{maximum of 10 points}. One student wrote only ``the factors that made Siam
break Longvek: the loss of solidarity'', naming a single correct cause. The teacher awarded
\textbf{2 out of 10}. The grading system must read both texts, recognise that the answer is partially
correct, and reproduce that partial mark. We turn the raw score into a scale-free target by dividing
by the maximum, $y = 2/10 = 0.2$, and into a five-class grade by rounding, $\ell =
\mathrm{round}(4 \times 0.2) = 1$. Every model in this study is trained to predict $y$ for inputs of
exactly this form, and is judged both on the grade it implies and on the raw mark it recovers.

\section{Aim and Objectives of the Study}
In this study, we aim to build the first reproducible, \emph{explainable} ASAG benchmark for Khmer and
to describe honestly what current model families can and cannot do on a real classroom corpus. To
achieve this aim, the objectives of this study are as the following:

\begin{itemize}[leftmargin=1.4em]
  \item To build and document a Khmer short-answer corpus with two cleaning variants and a
  Khmer-aware preprocessing pipeline.
  \item To implement one shared training-and-evaluation pipeline that compares four model families
  (classical, RNN, Transformer encoder, and a fine-tuned LLM) as bounded ordinal regression.
  \item To produce a model-agnostic SHAP word-attribution explanation and check its plausibility,
  applied to all four families.
  \item To evaluate every model with field-standard metrics across two dataset variants.
  \item To deliver the explainability feature as a live, teacher-facing web prototype.
\end{itemize}

These objectives sharpen into three research questions that the rest of the thesis answers in order:

\begin{itemize}[leftmargin=1.4em]
  \item \textbf{RQ1 (accuracy).} How well do the four model families agree with the human grader on
  Khmer short answers, measured by Quadratic Weighted Kappa, and does a more expensive model agree
  more closely?
  \item \textbf{RQ2 (deployment).} How often does each model reproduce the teacher's \emph{exact} mark
  in points, and is the most accurate model on QWK also the best for classroom use?
  \item \textbf{RQ3 (explainability).} Can one model-agnostic word-attribution method (SHAP) produce
  plausible explanations across all four families, including the non-differentiable classical model?
\end{itemize}

\section{Rationale of the Study}
This study is motivated by a clear gap: short-answer grading has advanced for high-resource languages
but not for Khmer, and a grade is a high-stakes decision that a teacher will only trust if it can be
\emph{audited}. An explanation is only useful if it points at the content that justified the grade, so
a teacher can check it against the reference answer~\citep{lyu2024faithfulsurvey}. We therefore treat
explainability as a first-class goal, using SHAP word attribution and checking that the words it
highlights are plausible (they match the reference content).

The study makes the following contributions:

\begin{enumerate}[leftmargin=1.8em]
  \item \textbf{The first Khmer ASAG benchmark}: 1{,}184 real teacher-graded answers (41 questions,
  203 students, four subjects, two schools), released with two cleaning variants.
  \item \textbf{A four-family comparison under one pipeline}: classical, RNN, Transformer, and a
  fine-tuned LLM (the KhmerGrader family).
  \item \textbf{A cross-family SHAP explanation}: one model-agnostic SHAP word-attribution method
  applied to all families, with a plausibility (reference-overlap) check.
  \item \textbf{A live teacher-facing prototype} that returns a score, a SHAP word-attribution
  explanation, and written feedback.
\end{enumerate}

To ground the data in reality, we collected real semester exam answers by hand from two Cambodian high
schools chosen to differ: one in a province and one in the capital. The two sites differ in students'
answering style and in the marking teacher, so a model that works across both is tested against real
classroom variation.

\section{Limitation and Scope}
This study covers short-answer questions in Khmer from four secondary-school subjects (Biology,
History, Geography, Earth Science). We study \emph{content} scoring against a reference answer, not
essay style or grammar. A single teacher graded every answer, so results are agreement \emph{with that
grader}, not against a measured human ceiling. The heavier models (the Transformer encoders and the
LLM) need a GPU and were run on the MISTI-HPC cluster (an NVIDIA A40); the classical and RNN baselines run on the CPU. We
revisit these limits in Chapter~\ref{ch:limitations}.

\section{Structure of the Study}
The rest of this thesis is organised as follows:

\begin{itemize}[leftmargin=1.4em]
  \item \textbf{Chapter~\ref{ch:literature}: Literature Review} reviews ASAG, the core concepts
  (Transformers and explainability), related work, and the research gap.
  \item \textbf{Chapter~\ref{ch:methodology}: Methodology} describes the research design, data
  collection, preprocessing, the four models, the tools, and the evaluation metrics.
  \item \textbf{Chapter~\ref{ch:experiments}: Experiment} describes the environment and the
  before-training, training, and testing phases, including the prototype.
  \item \textbf{Chapter~\ref{ch:results}: Results and Discussion} presents the results and discusses
  them.
  \item \textbf{Chapter~\ref{ch:limitations}: Limitations and Future Work} states the limits and the
  next steps.
  \item \textbf{Chapter~\ref{ch:conclusion}: Conclusion} concludes the study.
\end{itemize}
