\chapter{Literature Review}
\label{ch:literature}

This chapter reviews the work this study builds on: an overview of automatic short-answer grading
(\S\ref{sec:lit-asag}), the challenges of Khmer (\S\ref{sec:lit-khmer}), the core concepts of
Transformers and explainability (\S\ref{sec:lit-core}), related work (\S\ref{sec:lit-related}), and
the gap in existing research (\S\ref{sec:lit-gap}).

\section{Overview of Automatic Short-Answer Grading}
\label{sec:lit-asag}
Automatic Short-Answer Grading (ASAG) scores a free-text student answer by comparing it, directly or
indirectly, to a reference answer. The field has gone through four overlapping stages:

\begin{itemize}[leftmargin=1.4em]
  \item \textbf{Similarity-based.} Early systems scored an answer by how closely it matched the
  reference, using surface features such as word overlap, $n$-gram and edit distance, and later
  dependency-graph alignment between the parse trees of the answer and the reference~\citep{mohler2011}.
  These methods are transparent and need no training data, but they reward wording over meaning and miss
  paraphrases that use different words for the same idea.
  \item \textbf{Shared tasks.} Public datasets and common evaluation protocols, most influentially
  SemEval-2013 Task~7~\citep{dzikovska2013}, made systems comparable for the first time and introduced
  the key idea of testing on \emph{unseen questions} and \emph{unseen answers}.
  \item \textbf{Deep learning.} Work moved from hand-built features to learned representations: word
  embeddings, then recurrent networks, then Transformer encoders fine-tuned for scoring, often with data
  augmentation to cope with small corpora~\citep{sung2022survey,tan2025review}. These models capture
  meaning beyond surface overlap but need more data and are harder to interpret.
  \item \textbf{Large language models.} The current stage fine-tunes or prompts billion-parameter
  generative models to produce a score directly from the question, reference, and answer. The recurring
  finding is that a fine-tuned specialist model usually beats a larger general model used
  zero-shot~\citep{chang2024gpt4}, which is the design this study follows for its LLM pillar.
\end{itemize}

Throughout these stages, \emph{Quadratic Weighted Kappa} (QWK) has stayed the standard metric for
ordinal scoring, because it rewards near-misses and penalises large disagreements more than small ones;
it is usually reported alongside exact and adjacent (within-one-point) agreement so that a reader can
separate overall correlation from exact-mark accuracy.

\section{Challenges in Khmer}
\label{sec:lit-khmer}
Khmer is a low-resource language whose script makes grading harder than in English:

\begin{itemize}[leftmargin=1.4em]
  \item there are \textbf{no spaces between words}, so a system must first segment the text before any
  grading model sees it;
  \item the script stacks subscript consonants and reorders vowels inside a syllable, and it carries
  invisible characters that must be cleaned;
  \item multilingual models see little Khmer during pre-training.
\end{itemize}

Khmer NLP has focused on foundational tasks: word segmentation, part-of-speech tagging, and
pre-trained models~\citep{buoy2022khmer,khmernltk2020}. This study uses the \texttt{khmer-nltk}
segmenter for the segmentation step. More broadly, preparing Cambodian education for AI is an active
concern~\citep{huot2025cambodia}, but grading Khmer student answers is still unaddressed.

\section{Core Concepts: Transformers and Explainability}
\label{sec:lit-core}

\subsection{Transformers and Pre-trained Encoders}
The Transformer replaced recurrence with \emph{self-attention} and is the backbone of modern NLP.
Self-attention lets every token in a sentence look directly at every other token and weigh how much
each one matters for its own representation, so a model can link ``it'' to the noun it refers to, or an
answer phrase to the part of the reference it echoes, without passing information step by step through a
chain of hidden states the way a recurrent network must. Because every pair of positions is connected in
one layer, long-range dependencies are as easy to capture as short ones, and the computation
parallelises well on a GPU.

Pre-trained encoders built on this architecture learn general-purpose text representations from large
unlabelled corpora and can then be fine-tuned on a small labelled task. BERT and its multilingual
version mBERT~\citep{devlin2019bert} are trained with a masked-language-modelling objective (predict
hidden words from their context); cross-lingual encoders such as XLM-R extend this to a hundred
languages in one shared vocabulary, which is what makes them usable for a low-resource language like
Khmer that has too little text to pre-train a dedicated model. The limitation is that such models see
only a thin slice of Khmer during pre-training, so their Khmer representations are weaker than their
English ones. BERT, mBERT, XLM-R, GTE-multilingual, and the Qwen LLM used in this study are \emph{all}
Transformers; they differ in size, training objective, and how they are adapted, not in their core
self-attention design.

\subsection{Explainability}
Explainable AI (XAI) makes model decisions transparent. For text scoring the most useful family is
\emph{attribution}: assign each input word an importance score so a teacher can see which words drove
the grade. Attribution methods fall into three groups. \emph{Perturbation} methods change the input and
watch the output: occlusion (leave-one-out) removes one word at a time, while LIME fits a simple local
model to many random perturbations. \emph{Gradient} methods (plain saliency, Integrated Gradients) read
the model's gradients with respect to each input, so they need a differentiable model and access to its
internals. \emph{SHAP} (SHapley Additive exPlanations) distributes credit among the input words using
Shapley values, the average marginal contribution of a word over all coalitions of the
others~\citep{lundberg2017shap}; it is the field standard for explainable scoring and, in its
model-agnostic form, treats the model as a black box, so one method works for a neural network, a
Transformer, and the non-differentiable support-vector regressor alike.

This study uses SHAP as its explanation, following the explainable-scoring work of Kumar and
Boulanger~\citep{kumar2020explainable}, who use SHAP to expose which features drive an essay score and
show that such explanations carry genuine pedagogical value. We judge an explanation by its
\emph{plausibility}~\citep{lyu2024faithfulsurvey}: whether the words it highlights look right to a
teacher, measured as the overlap between the top SHAP words and the content of the reference answer.
A plausible explanation highlights the rubric-relevant content a teacher would check, which is what
turns an attribution into a basis for formative feedback.

\section{Related Work}
\label{sec:lit-related}
In grading specifically, XAI is used both to explain scores and to give feedback. Kumar and Boulanger
showed that explainable automated essay scoring has pedagogical value, using SHAP over interpretable
features~\citep{kumar2020explainable}. Most later scoring-XAI work has also relied on SHAP, LIME, or
Integrated Gradients~\citep{lachana2025dutch}, and one study reports that SHAP and LIME are poorly
aligned with rubric assessment~\citep{xu2025qwenscore}. A systematic comparison of attribution methods
for ASAG situates our choice of SHAP~\citep{pinto2025attribution}.
For evaluation, recent work compares fine-tuned models against prompted LLMs for grading and reports
QWK with exact and adjacent agreement~\citep{lachana2025dutch}. The closest comparable dataset is an
Arabic ASAG corpus on a three-point scale, whose best Transformer reaches 95.67\,\%/77.22\,\%
train/test accuracy and an unweighted Cohen's $\kappa$ of 0.60~\citep{alaoui2024}; we use it as the
anchor for honestly reporting the train/test gap.

Table~\ref{tab:priorwork} places this study against representative prior work, and the rows trace the
arc of the field. The early baselines establish the task and its metrics: Mohler et
al.~\citep{mohler2011} score answers by semantic similarity and dependency-graph alignment, and the
SemEval-2013 shared task~\citep{dzikovska2013} standardises evaluation and the unseen-question setting.
The survey of Sung et al.~\citep{sung2022survey} documents the shift from word embeddings to
Transformers. On the explainability side, Kumar and Boulanger~\citep{kumar2020explainable} show that
explainable essay scoring has genuine pedagogical value but rely on SHAP over hand-built features, and
Xu et al.~\citep{xu2025qwenscore} report that SHAP and LIME align only weakly with how a rubric is
actually applied. Pinto and Shin~\citep{pinto2025attribution} compare attribution methods for ASAG
directly, including SHAP and leave-one-out, informing our use of SHAP. On the model side,
Chang and Ginter~\citep{chang2024gpt4} test GPT-4 prompting and find fine-tuned specialists tend to
lead, and Lachana~\citep{lachana2025dutch} confirms this for Dutch ordinal scoring while reporting QWK
with exact and adjacent agreement, the same metric suite we adopt. Latif and
Zhai~\citep{latif2024finetuning} show that fine-tuning a general LLM substantially outperforms its
prompted base for scoring; and Solano et al.~\citep{solano2025narrowing} report that supervised
fine-tuning of \emph{open-source} LLMs narrows the gap to proprietary models for educational tools,
scoring with greedy decoding (temperature 0) and reasoning disabled, the same inference setting we use.
We extend this contrast to a low-resource language by comparing the
fine-tuned KhmerGrader against contemporary frontier models prompted zero-shot. For Khmer specifically, Buoy et
al.~\citep{buoy2022khmer} provide pre-trained models and evaluation data but no grading task. Metric
values are not comparable across rows (different datasets, languages, and scales); the table is a map of
\emph{approaches}, not a leaderboard.

{\footnotesize\setstretch{1}\setlength{\tabcolsep}{4pt}\renewcommand{\arraystretch}{1.2}
\begin{longtable}{@{}>{\raggedright\arraybackslash}p{4.2cm} >{\raggedright\arraybackslash}p{1.7cm} >{\raggedright\arraybackslash}p{2.6cm} >{\raggedright\arraybackslash}p{2.3cm} >{\raggedright\arraybackslash}p{2.7cm}@{}}
\caption[Related work in automated short-answer grading]{Related work in automated short-answer grading and the gap this study fills. Metric values are not comparable across rows.}\label{tab:priorwork}\\
\toprule
Paper (author, year) and title & Language / task & Model \& approach & XAI method & Headline result \\
\midrule
\endfirsthead
\multicolumn{5}{@{}l}{\itshape Table~\ref{tab:priorwork} continued from previous page}\\
\toprule
Paper (author, year) and title & Language / task & Model \& approach & XAI method & Headline result \\
\midrule
\endhead
\midrule
\multicolumn{5}{r@{}}{\itshape continued on next page}\\
\endfoot
\bottomrule
\endlastfoot
Mohler et al.~(2011)~\citep{mohler2011}\newline \emph{Learning to grade short answer questions using semantic similarity measures and dependency graph alignments} & English, short answer & Lexical similarity + dependency-graph alignment & n/a & Pearson correlation / RMSE (early baseline) \\
\addlinespace
Dzikovska et al.~(2013)~\citep{dzikovska2013}\newline \emph{SemEval-2013 Task 7: joint student response analysis and recognizing textual entailment} & English, SemEval-2013 Task~7 & Textual-entailment classification (shared task) & n/a & macro-F1 (standard benchmark) \\
\addlinespace
Kumar \& Boulanger~(2020)~\citep{kumar2020explainable}\newline \emph{Explainable automated essay scoring: deep learning really has pedagogical value} & English, essays (AES) & Deep-learning essay scoring & SHAP over features & QWK; shows XAI has pedagogical value \\
\addlinespace
Sung et al.~(2022)~\citep{sung2022survey}\newline \emph{A survey on automated short answer grading with deep learning: from word embeddings to transformers} & ASAG survey & Word embeddings $\to$ Transformers & n/a & survey (no single metric) \\
\addlinespace
Chang \& Ginter~(2024)~\citep{chang2024gpt4}\newline \emph{Performance of the large language model GPT-4 on automated short answer grading} & English, short answer & GPT-4 prompting (zero/few-shot) & n/a & QWK; fine-tuned specialists tend to lead \\
\addlinespace
Soulimani et al.~(2024)~\citep{alaoui2024}\newline \emph{Deep learning based Arabic short answer grading in serious games} & Arabic, 3-class & Deep learning (Transformer) & n/a & Acc 95.67/77.22 (train/test); Cohen $\kappa$ 0.60 \\
\addlinespace
Pinto \& Shin~(2025)~\citep{pinto2025attribution}\newline \emph{Evaluating the consistency and reliability of attribution methods in ASAG systems} & English, ASAG & Attribution-method comparison & LOO, SHAP, IG & consistency / reliability of attributions \\
\addlinespace
Xu et al.~(2025)~\citep{xu2025qwenscore}\newline \emph{Explainable AI for education: enhancing essay scoring via rubric-aligned chain-of-thought prompting} & Essays (LLM) & Rubric-aligned chain-of-thought prompting & SHAP, LIME & rubric-aligned scoring; SHAP/LIME weakly aligned \\
\addlinespace
Lachana~(2025)~\citep{lachana2025dutch}\newline \emph{Automated scoring systems for open-ended questions in Dutch education} & Dutch, ordinal & Fine-tuned vs prompted LLM & n/a & QWK, exact/adjacent; fine-tuned wins \\
\addlinespace
Buoy et al.~(2022)~\citep{buoy2022khmer}\newline \emph{Pretrained models and evaluation data for the Khmer language} & Khmer NLP & Pretrained models + evaluation data & n/a & foundational Khmer models (no grading) \\
\addlinespace
\textbf{This study}\newline \emph{Explainable AI for automated grading of Khmer language short-answer questions in education} & \textbf{Khmer, 5-class} & Classical SVR, BiLSTM, Transformer, LLM & SHAP & QWK 0.80--0.85; exact 66\% \\
\end{longtable}
}

As Table~\ref{tab:priorwork} shows, prior grading work spans several languages and model families, but
the XAI column is sparse: most grading work ships no word-level explanation at all, the explainable
studies apply SHAP to a single model type, and none targets Khmer. That is the gap this study fills.

\section{Gap in Existing Research}
\label{sec:lit-gap}
Two gaps stand out:

\begin{enumerate}[leftmargin=1.8em]
  \item there is \textbf{no published ASAG benchmark for Khmer}, so the field has no data and no
  baselines;
  \item most ASAG work ships no word-level explanation, and the explainable studies apply SHAP to a
  single model type rather than across diverse model families.
\end{enumerate}

This study addresses both: it builds the first reproducible Khmer ASAG benchmark and applies one
model-agnostic SHAP explanation across four diverse model families (including the non-differentiable
classical model).
