\chapter{Experiments and System Implementation}
\label{ch:experiments}

\section{Environment and Setup}
The pipeline is implemented in Python. The classical and recurrent models run on a CPU; the
Transformer encoders and the QLoRA-fine-tuned LLM require a CUDA GPU and were run on dedicated
hardware. Khmer word segmentation uses \texttt{khmer-nltk}. Every run writes a self-contained
directory holding its configuration, per-answer predictions for the train/validation/test splits,
and a metrics file, so that all reported numbers can be recomputed from saved predictions without
re-training. A fixed random seed (42) is used throughout, with multi-seed runs (five seeds) for the
variance-sensitive leakage analysis.

\section{The Experiment Grid}
\label{sec:grid}
The study is organised as a sequence of experiments, each isolating one design choice, plus the
explainability and statistical-rigour studies (Table~\ref{tab:grid}). Each modelling experiment runs
over the three dataset variants and the preprocessing/input-format axes, and writes its own
leaderboard.

\begin{table}[t]
\centering
\caption{The experiment grid. Each row isolates one design decision or analysis.}
\label{tab:grid}
\small
\begin{tabular}{@{}llp{8.0cm}@{}}
\toprule
ID & Pillar / study & What it adds \\
\midrule
exp01  & Classical    & TF-IDF cosine and RBF-SVR baselines \\
exp02  & Classical    & Post-hoc threshold calibration (validation-selected) \\
exp03  & Classical    & Max-score as an SVR feature \\
exp03b & Neural       & Max-score feature for neural heads \\
exp04  & Classical    & Per-max-score ``bucket'' SVR \\
exp05  & RNN          & Character BiLSTM + attention grid \\
exp06  & Transformer  & Dual / cross encoders $\times$ mBERT / XLM-R / GTE \\
exp07  & Ensemble     & Weighted top-3 cells by validation QWK \\
exp08  & LLM          & Qwen QLoRA fine-tuning \\
exp09  & Explainability & LOO faithfulness + plausibility across families \\
exp10  & Statistics   & Champion point metrics + question-leakage comparison \\
exp11  & Ablation     & Old vs.\ refined cleaning (classical champion) \\
exp12  & Tuning       & Classical SVR $C$ and TF-IDF max-features sweep \\
\bottomrule
\end{tabular}
\end{table}

\section{Training and Evaluation Protocol}
For every cell, models are trained on the training split and selected on the \emph{validation} split;
the chosen configuration is then reported \emph{once} on the held-out test split, with no test-set
peeking. Neural and Transformer models use early stopping on validation QWK. All metrics are
computed afterwards from the saved test predictions (\S\ref{sec:metrics}).

\section{Hyperparameter Tuning}
\label{sec:tuning}
Hyperparameters are selected by validation QWK. For the classical champion, which trains fast enough
to sweep exhaustively on a CPU, a grid over the SVR penalty $C\in\{0.1,1,10\}$ and the TF-IDF
maximum features $\{5,15,30\}\times10^{3}$ was run; the validation-selected configuration changed the
\emph{test} QWK by less than 0.002 relative to the defaults, indicating the classical result is not
sensitive to these choices. For the neural, Transformer, and LLM pillars, the analogous
learning-rate sweeps (BiLSTM $\in\{5\mathrm{e}{-}4,1\mathrm{e}{-}3,2\mathrm{e}{-}3\}$; Transformers
$\in\{1\mathrm{e}{-}5,2\mathrm{e}{-}5,5\mathrm{e}{-}5\}$; LLM QLoRA
$\in\{1\mathrm{e}{-}4,2\mathrm{e}{-}4,3\mathrm{e}{-}4\}$) use the same validation-selection rule and
are run on GPU hardware.

\section{System: the Live Teacher-Facing Prototype}
\label{sec:prototype}
The explainability contribution is realised as a working, hosted web application rather than only a
benchmark. The interaction is simple and mirrors what a teacher already does:

\begin{enumerate}[leftmargin=1.8em]
  \item the teacher types (or pastes) the \textbf{question}, the \textbf{reference answer}, and the
  \textbf{student's answer};
  \item the teacher \textbf{chooses a model} from the four pillars (classical, RNN, Transformer, or
  LLM) via a selector;
  \item the system returns a \textbf{score}, a \textbf{word-attribution explanation} (a text-highlight
  heatmap of the words that drove the grade), and \textbf{written feedback}.
\end{enumerate}

Two design points connect the tool back to the research. First, the explanation the teacher sees is
the \emph{same} occlusion word attribution that the faithfulness study validates, so the XAI study is the
feature, not an add-on. Second, the prototype is positioned as a \emph{teacher-assist} tool with the
human in the loop, it supports a teacher's judgement rather than replacing it. The live application is
hosted at \href{[demo URL]}{[demo URL to insert]}; example screenshots of the input form and a
returned score with its Khmer word-attribution heatmap and feedback are shown in
Appendix~\ref{app:prototype}.

\section{Reproducibility}
All code, the dataset and its three variants, the random seeds, the per-cell leaderboards, the
champion prediction files, and the figures are released. Because every metric is computed
from saved predictions, every result in Chapter~\ref{ch:results} can be reproduced without
re-training the models. Large model weights (LLM adapters and encoder checkpoints) are not tracked in
the repository but are regenerable from the released training scripts.
