\chapter{Limitations and Future Work}
\label{ch:limitations}

\section{Limitations}
We state the limits of this study plainly, since doing so is part of an honest benchmark. Each
limitation below is tied to a concrete result so that a reader knows not just \emph{that} a caveat
exists but how much it could matter.

\begin{itemize}[leftmargin=1.4em]
  \item \textbf{Single grader, no inter-annotator agreement.} A second grader could not be obtained
  within the scope of this work, so there is no measured human agreement ceiling. All results are
  agreement \emph{with this teacher}, and may inherit that grader's biases. Reporting across three
  dataset variants partially mitigates, but does not remove, this limitation, and the 10C Biology case
  (\S\ref{sec:noisy}) shows that even a single grader is not perfectly self-consistent.
  \item \textbf{Small corpus and few questions.} With roughly 909 to 1{,}184 answers across only 41
  distinct questions, the corpus covers four subjects from two schools rather than a national sample,
  and some cells are very small: Earth Science in particular contributes only nine answers.
  \item \textbf{Partial cross-pillar comparability.} The pillars were evaluated on two dataset
  variants (909/1{,}184): the classical, RNN, and LLM champions share the \texttt{no10c} test set,
  while the encoder was selected on \texttt{full}, so the champion metrics are computed on slightly
  different test sizes. Cross-pillar QWK differences should therefore be read as indicative rather than
  head-to-head, which is also why we lean on the within-0.05 ``comparable'' band rather than declaring
  a single QWK winner.
  \item \textbf{Explainability scope.} SHAP plausibility is reported for all four pillars on the
  \texttt{no10c} split; the LLM uses a \emph{capped} SHAP evaluation budget, since each evaluation is a
  full generation, so its plausibility is an approximation of the full sweep. The plausibility measure is
  itself an automatic reference-overlap proxy, not a human-rationale study, so it approximates rather
  than measures pedagogical usefulness.
  \item \textbf{Frontier baselines are a dated API snapshot.} The frontier-model comparison
  (\S\ref{sec:frontier}) calls closed, continuously-updated services, so those numbers are not exactly
  reproducible if a provider revises or retires a version. We pin each version, record the evaluation
  date, and cache the raw responses as mitigation; the KhmerGrader weights, code, and data remain fully
  reproducible.
  \item \textbf{No formal field study.} The live prototype demonstrates feasibility, but its classroom
  value has not yet been evaluated with teachers in a controlled study.
\end{itemize}

\section{Future Work}
\label{sec:future}
Several directions follow naturally and would address the limitations above.

\begin{itemize}[leftmargin=1.4em]
  \item \textbf{A second grader.} Adding a second annotator would give a true agreement ceiling, so
  that results could be reported against measured human agreement rather than against a single grader.
  \item \textbf{Few-shot prompting of the bases and frontier models.} We prompt only zero-shot; adding
  in-context examples (for instance one calibration exemplar per grade band) could strengthen the
  prompted baselines. The example count and selection matter as much as their presence, and prompting is
  known to be prone to majority-label and recency biases, so a careful few-shot study, including
  boundary-aware exemplar selection, is left as future work.
  \item \textbf{A larger, multi-school corpus.} Extending collection beyond the two schools and four
  subjects, with more distinct questions, would make the benchmark more representative of Cambodian
  secondary education.
  \item \textbf{Khmer-specific pretraining.} Continued pretraining of a multilingual encoder on Khmer
  text, or a Khmer-native encoder, may close the gap left by the limited Khmer seen during generic
  multilingual pretraining.
  \item \textbf{A human plausibility study.} The automatic reference-overlap plausibility proxy could be
  replaced with a human-rationale study, in which teachers mark the words they consider relevant and the
  SHAP attributions are scored against those rationales.
  \item \textbf{Cross-lingual transfer.} Transferring scoring ability from high-resource languages to
  Khmer (for example, via multilingual representations or cross-lingual contrastive learning) could
  help in the genuinely low-data regime.
  \item \textbf{Feedback quality.} The prototype's written feedback should be evaluated for
  pedagogical usefulness in classroom settings, building on rubric-aligned feedback work.
  \item \textbf{A teacher field study.} A controlled study of how Cambodian teachers use, trust, and
  benefit from the prototype would test the explainability contribution where it ultimately matters.
\end{itemize}

Of these, the highest priority is a second grader, because it attacks the result that most limits the
study's external validity: it would replace agreement-with-one-teacher with a measured human ceiling. A
larger, multi-school corpus with more distinct questions is the natural next step. The remaining
directions are valuable but build on top of those two foundations.
