\chapter{Conclusion}
\label{ch:conclusion}

\section{Conclusions}
\label{sec:conclusions}
In this study, we set out to build the first reproducible, \emph{explainable} automatic short-answer
grading benchmark for Khmer, and to describe honestly what current model families can achieve on it. We
delivered a 1{,}184-answer teacher-graded corpus of real semester examination answers from two
Cambodian high schools, with two curated variants, a Khmer-aware preprocessing pipeline, a unified
pipeline comparing four model families (classical, RNN, Transformer, and a fine-tuned LLM), a
cross-family SHAP explanation study, and a live teacher-facing prototype.

The main findings can be summarised as follows. On this real Khmer exam corpus the most expensive model
does \emph{not} win the research metric: all four pillars reach QWK in the 0.80--0.85 band, and a
30-second CPU classical model is within 0.05 QWK of a fine-tuned LLM. Data cleaning lifts the same model
by $+0.027$ QWK, at least as much as swapping architectures on this small corpus, with the single biggest
step being removal of the inconsistently graded 10C Biology subset. On QWK the pillars are comparable,
and the fine-tuned LLM is the clear winner only on the deployment metrics (66\,\% exact match, 83\,\%
within one point, MAE $\approx$0.93 points). Among the cheap additions, the max-score feature gives
consistent, near-free gains, while post-hoc calibration is fragile and model-dependent and so is reported
as an ablation rather than in the headline. SHAP word attribution provides a plausible word-level
explanation for every model family, following the explainable-scoring approach of Kumar and Boulanger:
the words it highlights are mostly content words that also appear in the reference answer, so a teacher
can audit the grade rather than take it on trust.

Read against the three research questions, the answers are now in hand. On RQ1 (accuracy), the four
pillars agree with the teacher about equally well, in a narrow 0.05 QWK band, so the most expensive model
does not win the research metric. On RQ2 (deployment), the fine-tuned LLM is the clear winner on the
exact mark (66\,\%) even though its QWK only ties the others, which is exactly why the two questions must
be kept apart. On RQ3 (explainability), one model-agnostic SHAP method produces plausible
word-level explanations for all four families, including the non-differentiable classical model,
highlighting the content words a teacher would check against the reference answer.

The central message is that for Khmer short-answer grading, accuracy, deployment quality, and
explainability must be judged \emph{together}: no single pillar dominates all three. A 30-second CPU
classical model is competitive on the research metric and explains itself through SHAP word
attribution; a fine-tuned LLM is the right choice when exact teacher scores matter; and, above all, an
explanation is only worth showing to a teacher if its highlighted words are plausible against the reference.

The broader contribution is to Cambodian educational NLP, which has had no grading benchmark to build on.
By releasing the dataset and its two variants, the full code and seeds, the per-cell results, and the
KhmerGrader models, and by realising the work as a usable teacher-facing prototype, this thesis turns a
gap into a starting point: a reproducible baseline that others can extend, correct, and beat. It also
leaves a reusable methodological lesson that outlasts the specific numbers, that grading systems should
be evaluated on all three axes at once, and reported honestly as agreement with a single human grader
rather than against an unmeasured ceiling, so that the field reports what its models can really do.
